Final Project

### 2025-04-30 MSSC5750_Project.qmd
Added Shap Values & more Viz code

```{r echo = FALSE, message=FALSE}
library(MASS)
library(ggplot2)
library(dplyr)
library(scales)
library(glmnet)
library(fastDummies)
library(caret)

@@

@@

@@

>>>>

```{r echo = FALSE, message=FALSE}
library(MASS)
library(ggplot2)
library(dplyr)
library(scales)
library(glmnet)
library(fastDummies)
library(caret)
library(shapr)
library(shapviz)
library(party)
library(reshape2)

@@

# Oddly higher credit score does not seem to make a difference
cor(data$credit_score,data$loan_status)
summary(data$credit_score)
lowcred<-data %>% filter(credit_score<601)
goodcred<-data %>% filter(credit_score>670)
mean(lowcred$loan_status)
mean(goodcred$loan_status)

@@

temp<-data %>% summarize('0'=1-mean(loan_status),"1"=mean(loan_status) ,.by=loan_intent)
intent<-melt(temp) %>% dplyr::rename(loan_status=variable)

ggplot(data=intent, aes(x=loan_intent, y=value, group=loan_status, fill=loan_status)) + geom_col(position='stack')

@@

Shap values/viz
```{r}
# Requires split out xtrain, ytrain, xtest... Also a combined x/ytrain

# We will apply to lasso data

# We need to manually generate the dummy variables and match naems to what the model created
# coder<-dummyVars(~ person_home_ownership + loan_intent + previous_loan_defaults_on_file, data = data, fullRank = TRUE, sep='')
# trainDataAdj<-cbind(trainData %>% select(-person_home_ownership, -loan_intent, -previous_loan_defaults_on_file), predict(coder, newdata = trainData) )
# testDataAdj<-cbind(testData %>% select(-person_home_ownership, -loan_intent, -previous_loan_defaults_on_file), predict(coder, newdata = testData) )

# Only want the columns that we included in lasso

xtrain<-trainData %>% select(person_home_ownership ,loan_amnt , loan_intent , loan_int_rate , loan_percent_income  , credit_score , previous_loan_defaults_on_file)
ytrain<-trainData %>% select(loan_status)
xytrain<-trainData %>% select(person_home_ownership ,loan_amnt , loan_intent , loan_int_rate , loan_percent_income  , credit_score , previous_loan_defaults_on_file,loan_status)
xtest<-testData  %>% select(person_home_ownership ,loan_amnt , loan_intent , loan_int_rate , loan_percent_income  , credit_score , previous_loan_defaults_on_file)

# calculating shap is costly so we will reduce the size of the data we pass in

# we will just explain on our train data as well for larger sample size
set.seed(5)
xtrainSmall <- xtrain[sample(nrow(xtrain), 1000), ]
explainer<- shapr::explain(model=logModelLasso, x_train=xtrainSmall, x_explain=xtrainSmall, approach="ctree", phi0=mean(ytrain$loan_status))

shp <- shapviz(explainer)
sv_importance(shp)

plot(explainer, plot_phi0 = FALSE, index_x_explain = 1)

shapObj <-shapviz(explainer)
sv_importance(shapObj, kind="beeswarm", show_numbers = TRUE, max_display=15)
